```{r, include=FALSE}
#---
#title: "Chapter 5 analysis"
#author: "Antti Kaihovaara"
#date: "December 8, 2018"
#description: "Chapter 6 analysis"
#Data1: https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt
#Data2:https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt
#---
```

#Exercise 6

In this exercise we use two longitudinal datasets, RATS and BPRSL. Both datasets have been converted to long form for analyses.

We will introduce the datasets later, but first we have to convert the categorical variables in the data as factors. 

```{r, include=FALSE} 
getwd()

BPRSL <- read.csv("BPRSLong.csv", sep=",", header=TRUE)
RATSL <- read.csv("RATSLong.csv", sep=",", header=TRUE)

library(tidyverse)
library(dplyr)
library(tidyr)
library(ggplot2)
```


```{r, include=TRUE} 
BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)
glimpse(BPRSL)

RATSL$ID <- factor(RATSL$ID)
RATSL$Group <-factor(RATSL$Group)
glimpse(RATSL)

```


##RATS Longitudinal data

The RATS dataset includes information on nutrition of rats during 9 weeks period. There are three different groups and the key variable to follow is the weight of the rats.

The figure 1 below shows the development of rats' weights over 9 weeks in three different groups. Based on this graph it seems like the weight rose most in the groups 2 and 3.

```{r, include=TRUE} 
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))
```

As the groups start from different levels, it makes sense to standardize the values of each observation. Figure 2 below reveals that there is not too much differences between the groups in terms of weight changes.

```{r, include=TRUE} 
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdWeight = (Weight-mean(Weight))/sd(Weight)) %>%
  ungroup()


p1 <- ggplot(RATSL, aes(x = Time, y = stdWeight, linetype = ID))
p2 <- p1 + geom_line() + scale_linetype_manual(values = rep(1:10, times=4))
p3 <- p2 + facet_grid(. ~ Group, labeller = label_both)
p4 <- p3 + theme_bw() + theme(legend.position = "none")
p5 <- p4 + theme(panel.grid.minor.y = element_blank())
p6 <- p5 + scale_y_continuous(name = "standardized weight")
p6
```

There are quite many slopes in the two graphs above. Maybe we should look at the means of each group. This graph tells pretty much the same story as the previous one. It looks like the rats in group 2 gained most weight on average, but differences are small and some confidence intervals are overlapping.

```{r, include=TRUE} 
n <- length(0:8) # Time, incl. baseline (0)
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean=mean(Weight), se=sd(Weight)/sqrt(n) ) %>%
  ungroup()
glimpse(RATSS)

p1 <- ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group))
p2 <- p1 + geom_line() + scale_linetype_manual(values = c(1,2,3))
p3 <- p2 + geom_point(size=3) + scale_shape_manual(values = c(1,2,3))
p4 <- p3 + geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3)
p5 <- p4 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p6 <- p5 + theme(legend.position = c(0.5,0.5))
p7 <- p6 + scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
p7
```

In the figure 8.4 we have even more detailed graph on the observations on each group. This figure suggests that there might be some outliers in the data as some of the observations are quite far away from the boxes.

```{r Figure 8.4, echo=TRUE}
p1 <- ggplot(RATSL, aes(x = factor(Time), y = Weight, fill = Group))
p2 <- p1 + geom_boxplot(position = position_dodge(width = 0.9))
p3 <- p2 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p4 <- p3 + theme(legend.position = c(0.9,0.4))
p5 <- p4 + scale_x_discrete(name = "Time")
# Black & White version:
#p6 <- p5 + scale_fill_grey(start = 0.5, end = 1)
p5
```

One handy way of describing the data is to look at the mean boxplots in each group as is done in Figure 8.5. However, this figure makes the time dimension to disappear. We can make the same observation as in the previous graph: we do have some outliers in the data.

```{r Figure 8.5, echo=TRUE}

RATSL8S <- RATSL %>%
  filter(Time> 0) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()
 glimpse(RATSL8S)


p1 <- ggplot(RATSL8S, aes(x = Group, y = mean))
p2 <- p1 + geom_boxplot()
p3 <- p2 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p4 <- p3 + stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white")
p5 <- p4 + scale_y_continuous(name = "mean(Weight), Time 1-60")
p5
```
  
    
      

Removing the outliers narrows the confidence intervals and makes our measurement more accurate. However, the big picture remains the same.

```{r Figure 8.6, echo=TRUE}
# Remove the outlier:
RATSL8S1 <- RATSL8S %>%
  filter(mean < 550)

RATSL8S2 <- RATSL8S1 %>%
  filter(mean > 250)

 p1 <- ggplot(RATSL8S2, aes(x = Group, y = mean, outline=FALSE))
p2 <- p1 + geom_boxplot(outline=FALSE)
p3 <- p2 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p4 <- p3 + stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white")
p5 <- p4 + scale_y_continuous(name = "mean(Weight), Time 1-60")
p5
```

Finally, we run a one-way analysis of variance (ANOVA) to see whether the differences between the groups are statistically significant. To do this, we add the baseline, which is the starting week of the diet (WD1). The baseline values are strongly related tothe values that are taken after the treatment has begun, which is what we would expect. More importantly, it seems that the differences between groups are also significant at the 95 % confidence level. (I had some problems knitting the code, so I just copypasted the table. Keep calm and use Stata.)



Analysis of Variance Table

Response: mean
          Df Sum Sq Mean Sq   F value    Pr(>F)     
baseline   1 252125  252125 2237.0655 5.217e-15 ***  
Group      2    726     363    3.2219   0.07586 .    
Residuals 12   1352     113                        
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1




##BPRS Longitudinal data

The BPRS dataset includes 40 males, who were randomly assigned to control group and treatment group. Each subject was rated on the brief psychiatric rating scale (BPRS) before the treatment and after that each week for eight weeks. The idea of the scale is to evaluate schizophrenia among the patiens.

First, let's look at the within-subject scales changed in the two groups in 8 weeks.

```{r, include=TRUE} 
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
``` 

The previous graph did not reveal too much about the differences between the groups. Next we will run a naive OLS model using bprs as depent variable and weeks and treatment as independent variables. Based on this, it looks like treatment does not have a statistically significant association, but weeks do. However, this model assumes that the observations are independent from each other, which is rarely the case in longitudinal data.

```{r, include=TRUE} 
BPRS_reg <- lm(bprs ~ week + treatment, data = BPRSL)

summary(BPRS_reg)
```

Fitting a random intercept model takes into account the fact that our observations are not independent. It lets our subjects' intercepts to differ from other subjects intercepts. The results look still very similiar to simple OLS model.

```{r, include=TRUE} 
library(lme4)

BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)

# Print the summary of the model
summary(BPRS_ref)
``` 

Next, we fit a random intercept and random slope model, which allows differences in intercepts and in slopes. The advantage of this approach in comparison to the previous model is that this model takes into account the time dimension: it allows us to look at the differences in how bprs changes between and within subjects over time. Yet again, we do not see too many changes. The coefficient for time (week) is significant, but the coefficient for treatment is not. 

Analysis of variance shows us that the model with random intercepts and slopes (BPRS_ref1) is a better fit from our two random effects models.
 
```{r, include=TRUE}
BPRS_ref1 <- lmer(bprs ~ week + treatment+ (week | subject), data = BPRSL, REML = FALSE)

summary(BPRS_ref1)

anova(BPRS_ref1, BPRS_ref)
``` 

Adding interaction term (week*treatment) in the model should show us whether the time related effects are different for those who are treated and those who are not. It seems that this is not the case, as the the interaction effect is not statistically significant. However, the model including the interaction term seems to be the better fit from the two models.

```{r, include=TRUE}

BPRS_ref2 <- lmer(bprs ~ week * treatment + (week | subject), data = BPRSL, REML = FALSE)

summary(BPRS_ref2)

anova(BPRS_ref2, BPRS_ref1)

```

```{r, include=TRUE}

ggplot(BPRSL, aes(x = week, y = bprs, group = subject)) +
  geom_line(aes(linetype = treatment)) +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))

Fitted <- fitted(BPRS_ref2)

BPRSL <- BPRSL %>%
  mutate(Fitted)


ggplot(BPRSL, aes(x = week, y = Fitted, group = subject)) +
  geom_line(aes(linetype = treatment)) +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))

```

Finally, we plot the observed changes in 8 weeks period in treatment group and the control group (same graph as in the beginning) and plot the fitted model (the one with interactions). Just by looking at the graphs, we can tell that our model is actually a pretty good fit as the fitted slopes tell the same story as the graph based on observations. Furthermore, the fitted graph shows that the slope of bprs scores declined a little bit more in group "treatment1"" than in "treatment2". However, out regressions suggest that this difference is not statistically significant. 


